---
title: "SpaNorm clean: sample.p = 0.25"
subtitle: "clusters 15 + 20 removed"
date: last-modified
---

## Why this notebook

The  SpaNorm build at `sample.p = 0.25` (`20260513_SpaNorm_samplep025.qmd`) yields 25 clusters at Louvain 0.5. Two of them, `seurat_clusters` 15 and 20, look like contamination on the per-slide UMAP and carry top markers we read as artefactual rather than biological:

- Cluster 15 (parent annotation: mixed reactive / disease-associated, likely doublets). Top markers: Serpina3n + Gpnmb + Car12 reactive triad mixed with the neuronal markers Grin2a / Grin2b / Gad1 / Dgat2 / Tnr. Read as plaque-microenvironment doublets between reactive microglia / astrocytes and their neuronal neighbors.
- Cluster 20 (parent annotation: excitatory neurons, Calb1+/Tdo2+ of uncertain region). Top markers: Calb1 + Tdo2 + Grin2a / Grin2b + Car12 / Dgat2 / Tnr. Tdo2 and Calb1 mark dentate gyrus, habenula, and cerebellar Purkinje / granule layers (Allen Brain Atlas, mouse), so the cluster is anchored on regionally restricted genes and biases by section coverage rather than by biology shared across the seven samples.

We re-fit per-sample SpaNorm at `sample.p = 0.25` on the raw merged object after dropping the cells assigned to those two clusters, and we re-cluster from PCA forward. 

Every other parameter is held fixed against the parent notebook so the before-after comparison is clean: same seven per-sample SpaNorm fits, same `sample.p = 0.25`, same `df.tps = 6`, same `ScaleData -> PCA(30) -> FindNeighbors -> Louvain -> UMAP` chain, same 480-gene variable-feature set. We bracket Louvain at 0.5 and 0.7 on the same PCA / SNN graph and stash each cluster-id vector as a named metadata column (`clusters_res05` / `clusters_res07`) on a single saved Seurat object, matching the structure of `20260517_microglia_subcluster.qmd`.


Bibliography for this notebook:

- Salim et al., "SpaNorm: spatially-aware normalization for spatial transcriptomics data," Genome Biology, 2025. DOI: 10.1186/s13059-025-03565-y.
- SpaNorm package vignette: https://bhuvad.github.io/SpaNorm/articles/SpaNorm.html.
- Allen Brain Atlas, mouse, for Tdo2 / Calb1 regional expression: https://mouse.brain-map.org.
- Parent notebook: `20260513_SpaNorm_samplep025.qmd` (promoted project default).

## Setup

```{r}
#| label: setup

# Seurat v5 holds the merged Xenium object with the per-slide FOV slots we read centroids from.
library(Seurat)
library(tidyverse)
# qs2 reads the multi-GB checkpoint roughly 10x faster than saveRDS.
library(qs2)
library(fs)
# SpaNorm is the spatially aware normalization (Salim, Genome Biology, 2025).
library(SpaNorm)
# SpatialExperiment is the native input class for SpaNorm; we build one per sample.
library(SpatialExperiment)
# scCustomize only contributes its categorical palette helper for the UMAP plots.
library(scCustomize)
# kableExtra renders the per-cluster top-marker table inside a scroll box.
library(kableExtra)

# Source the shared helper functions; fiddle() draws the per-cluster violin panel.
source("R/20260514-helper_functions.R")

# Seed Louvain and any other stochastic step so cluster labels stay stable across renders.
set.seed(1)

# Reload the clean SpaNorm checkpoint; the build chunk below writes both clusters_res05
# and clusters_res07 as named metadata columns on the same object.
xen <- qs_read("data/20260516_microglia_switch_spanorm_p025_clean.qs2")
```

## Build the clean SpaNorm assay (drop clusters 15 + 20, re-fit per sample, bracket Louvain 0.5 / 0.7)

```{r}
#| label: spanorm-clean
#| eval: false


# Sample / slide / condition manifest; matches R/20260512_xenium2seurat.R for self-containment.
samples <- tribble(
  ~sample_id, ~slide,            ~condition,
  "4s2_F1",   "Slide1_0022474",  "control",
  "4s2M_F1",  "Slide1_0022474",  "switched",
  "4s2M_F2",  "Slide1_0022474",  "switched",
  "4s2_F2",   "Slide2_0069080",  "control",
  "4s2_F3",   "Slide2_0069080",  "control",
  "4s2M_F3",  "Slide2_0069080",  "switched",
  "4s2M_F4",  "Slide2_0069080",  "switched")

# Start from the promoted parent build and drop the two contamination clusters before re-fitting.
# SpaNorm needs raw counts, so we strip the object back to the Xenium assay and drop the stale
# SpaNorm normalization, PCA / UMAP, and seurat_clusters before re-fitting.
xen <- qs_read("data/20260512_microglia_switch_spanorm_p025.qs2")
DefaultAssay(xen) <- "Xenium"

# Confirm both cluster levels are present before subsetting; the previous version of this
# notebook silently no-oped on a non-existent "25" level (gotcha logged in CLAUDE.md).
stopifnot(all(c("15", "20") %in% levels(xen$seurat_clusters)))
xen <- subset(xen, subset = seurat_clusters %in% c("15", "20"), invert = TRUE)

xen <- DietSeurat(xen, assays = "Xenium", dimreducs = NULL, graphs = NULL)

# Literal slide -> FOV map; the merge of the two LoadXenium slides kept the FOVs as fov / fov.2.
slide_to_fov <- c(Slide1_0022474 = "fov", Slide2_0069080 = "fov.2")
fov_lookup   <- set_names(slide_to_fov[samples$slide], samples$sample_id)

# Run SpaNorm on one sample; SpaNorm's spatial smoothness assumption is per tissue section.
spanorm_one_sample <- function(s, xen, fov_lookup) {
  # One tissue section at a time; subsetting first keeps the TPS basis fitting only to that section.
  seu_s  <- subset(xen, subset = sample_id == s)
  # Centroids from the slide FOV; filter to the cells that survived the per-sample subset.
  coords <- GetTissueCoordinates(seu_s, image = fov_lookup[[s]]) |>
    filter(cell %in% Cells(seu_s))
  # SpatialExperiment is SpaNorm's native input; raw counts plus the 2D centroids.
  spe <- SpatialExperiment(
    assays        = list(counts = GetAssayData(seu_s, assay = "Xenium", layer = "counts")),
    spatialCoords = as.matrix(coords[, c("x", "y")]))
  # sample.p = 0.25 and df.tps = 6 held identical to the parent build for a like-for-like comparison.
  spe <- SpaNorm(spe,
                 sample.p   = 0.25,
                 df.tps     = 6,
                 adj.method = "auto",
                 backend    = "auto",
                 verbose    = TRUE)
  out <- SummarizedExperiment::assay(spe, "logcounts")
  colnames(out) <- Cells(seu_s)
  out}

# Run SpaNorm on each of the seven samples; map keeps the per-sample matrices named for the reduce step.
norm_list <- map(samples$sample_id, spanorm_one_sample,
                 xen = xen, fov_lookup = fov_lookup) |>
  set_names(samples$sample_id)

# Bind the per-sample matrices into one wide normalized matrix; reorder to the merged cell order.
# Qualify reduce() because SpatialExperiment loads IRanges, whose reduce() masks purrr::reduce.
norm_mat <- purrr::reduce(norm_list, cbind)
norm_mat <- norm_mat[, Cells(xen)]

# Attach as a v5 SpaNorm assay; data layer only because the auto-selected adjustment returns log-residuals.
# CreateAssay5Object avoids the "counts layer not present" warning that CreateAssayObject (v3) triggers.
xen[["SpaNorm"]] <- CreateAssay5Object(data = norm_mat)
DefaultAssay(xen) <- "SpaNorm"

# All 480 panel genes stay as variable features; FindVariableFeatures("vst") fails on a v5 assay
# carrying only the data layer, and the targeted Xenium panel is small enough that every gene
# is informative by design.
VariableFeatures(xen) <- rownames(xen)

# A v5 assay carrying only the data layer needs an explicit ScaleData before RunPCA.
xen <- ScaleData(xen, verbose = FALSE)

# 30 PCs match the parent build; default npcs = 50 is fine.
xen <- RunPCA(xen, npcs = 30, verbose = FALSE)

# Build the SNN graph once, then run Louvain at both resolutions inside the build.
# Each FindClusters call overwrites seurat_clusters and Idents(xen), so stash each result
# in a named metadata column; downstream chunks switch resolutions via group.by / Idents().
xen <- FindNeighbors(xen, reduction = "pca", dims = 1:30, verbose = FALSE)

xen <- FindClusters(xen, resolution = 0.5, algorithm = 1, verbose = FALSE)
xen$clusters_res05 <- xen$seurat_clusters

xen <- FindClusters(xen, resolution = 0.7, algorithm = 1, verbose = FALSE)
xen$clusters_res07 <- xen$seurat_clusters

xen <- RunUMAP(xen, reduction = "pca", dims = 1:30, verbose = FALSE)

# qs_save(xen, "data/20260516_microglia_switch_spanorm_p025_clean.qs2")
```

## Louvain at resolution 0.5

```{r}
#| label: palette-05
# Palette sized to the Louvain-0.5 cluster count; this section uses the clusters_res05 column.
colorines_05 <- scCustomize_Palette(num_groups = n_distinct(xen$clusters_res05), ggplot_default_colors = FALSE)
```

### PCA per slide

```{r}
#| label: pca-05
# Slide split so any residual slide-driven sub-structure surfaces before the UMAP.
PCAPlot(xen, group.by = "clusters_res05", split.by = "slide", cols = colorines_05)
```

**Result.** Both slides span the same PC1 / PC2 envelope and the major clusters sit in matched positions. 


### UMAP

```{r}
#| label: umap-05
# label = TRUE prints the cluster id inside each island so the marker panels below stay legible.
DimPlot(xen, group.by = "clusters_res05", cols = colorines_05, label = TRUE, label.size = 5) + NoLegend()
```



### Per slide

```{r}
#| label: umap-05-slide
#| fig-width: 10
#| column: page

# Side-by-side UMAPs per slide; failure mode is any residual cluster that lives on only one slide.
DimPlot(xen, group.by = "clusters_res05", split.by = "slide", cols = colorines_05) + NoLegend()
```

**Result.** Slide totals are Slide1 ≈ 140k cells vs Slide2 ≈ 277k.  
 The baseline Slide2 : Slide1 ratio is ~2:1 and any cluster ratio more than ~2x off that baseline could be slide-imbalanced. Two clusters stand out:

1. **Cluster 9 (mature OL with lower-FC values) is Slide1-specific:** 17,865 cells on Slide1 vs 468 on Slide2 (ratio 0.03:1; ~75x off the baseline). The marker set is identical to cluster 0 (Mag / Aspa / Mog / Gatm / Ermn / Cldn11) but the FCs collapse from ~3.7 to ~2.6-3.3. 

Proportionally the two slides actually carry similar OL fractions: Slide1 (0 + 9) = 25,596 / ~140k cells = 18.3 % and Slide2 (0 + 9) = 45,832 / ~277k cells = 16.5 %. The Slide1 OL pool splits between a high-FC core (cluster 0) and a lower-FC twin (cluster 9), while the Slide2 OL pool concentrates almost entirely in the high-FC cluster. Two candidate explanations remain open at this point in the analysis:

  * **Technical (normalization residual).** Slide2 carries +25 % transcripts / cell vs Slide1. SpaNorm corrects per section but is not guaranteed to fully equalize a transcript-yield gap of this size. 
  * **Biological (regional or maturation state).** The two FC tiers could reflect different anatomical OL pools (e.g. corpus callosum vs cerebellar / brainstem white matter) sampled differently across the two slides, or two maturation states of mature OLs.

2. **Cluster 23 (low-FC multi-lineage doublets, Slc32a1+/Mbp+/Cldn11+) is almost pure Slide2:** 7 cells on Slide1 vs 993 on Slide2 (ratio 142:1). Top markers are inhibitory-vesicular (Slc32a1) and myelin (Mbp / Cldn11 / Mog) at FCs ≤ 1.1; this is a small segmentation-doublet cluster between GABAergic neurons and adjacent OLs that materialises only where the GABAergic / white-matter interface is sampled densely (Slide2 sections 4s2_F2 and 4s2_F3 carry it).

3. **Five further clusters carry a moderate Slide2 over-representation** (cluster 13 Pou3f1+ stressed neurons at 4.66:1, cluster 24 T cells at 4.66:1, cluster 2 microglia at 3.07:1, cluster 19 Tshz2+ deep/hippocampal at 3.19:1, cluster 0 OL high-FC at 5.87:1) but they all reach > 90 cells on Slide1 and are not slide-private. 

4. The remaining 18 clusters sit within a 1.5-2.5x window of the 2:1 baseline.

### Per sample

```{r}
#| label: umap-05-sample
#| fig-width: 14
#| fig-height: 12
#| column: page

# Per-sample facets; 4s2M_F2 (partial section) still expected to contribute fewer cells than the others.
DimPlot(xen, group.by = "clusters_res05", split.by = "sample_id", ncol = 3, cols = colorines_05) + NoLegend()
```

**Result.** Six of the seven samples reproduce the 25-cluster landscape with comparable proportions; 4s2M_F2 is sparser everywhere, as expected for the partial section. Three sample-level patterns warrant flagging because they explain the slide imbalances above:

- **Cluster 9 (Slide1 OL twin) lives almost entirely in three samples:** 4s2_F1 (6,760), 4s2M_F1 (6,697) and 4s2M_F2 (4,408). The four Slide2 samples contribute ≤ 213 cells each. This is the cell-count breakdown behind the slide-level cluster-9 split; whether it reflects regional sampling or a transcript-yield residual remains open (see Per slide above).
- **Cluster 23 (Slide2 doublet cluster) lives in 4s2_F2 (493) and 4s2_F3 (262)** and is absent or near-absent everywhere else. A two-sample artefact, not a slide-wide phenomenon.
- **Cluster 13 (Pou3f1+ stressed neurons) is dominated by 4s2M_F3 (4,272)** and contributes < 2,300 cells in every other complete section. The Slide2 over-representation of this cluster is mostly one animal, not a slide-wide pattern.

Everything else (microglia, astrocytes, endothelial, neuronal subclasses, T cells) populates every complete section and balances across the genotype contrast.

### Find all markers

```{r}
#| label: find-all-markers-05
#| eval: false

# devtools::install_github('immunogenomics/presto')
# presto gives a far faster Wilcoxon Rank Sum implementation than Seurat's default.
library(presto)

# FindAllMarkers needs cluster identities as the active Idents(); set from the named column.
Idents(xen) <- "clusters_res05"
spanorm_markers_05 <- FindAllMarkers(xen, only.pos = TRUE, densify = TRUE)

write_csv(spanorm_markers_05, file = "results/20260516-find_all_markers_spanorm_p025_clean_res05.csv")
```

```{r}
#| label: markers-table-05

# Reload the cached marker table so this chunk renders without rerunning FindAllMarkers.
spanorm_markers_05 <- read_csv("results/20260516-find_all_markers_spanorm_p025_clean_res05.csv")

# Top 10 markers per cluster, sorted by avg_log2FC within cluster.
spanorm_markers_05 <- spanorm_markers_05 |>
  select(cluster, gene, avg_log2FC, p_val_adj) |>
  group_by(cluster) |>
  arrange(desc(avg_log2FC), .by_group = TRUE)

kbl(spanorm_markers_05 |> slice_head(n = 10)) |>
  kable_styling("striped") |>
  scroll_box(height = "500px")
```

```{r}
#| label: markers-violin-05
#| column: page
#| fig-height: 25
#| fig-width: 12

# Top 5 markers per cluster feed fiddle(); the SpaNorm assay carries the log-scale normalized values.
top_5_genes_05 <- spanorm_markers_05 |>
  slice_head(n = 5) |>
  pull(gene) |>
  unique()

fiddle(xen,
       genes_of_interest = top_5_genes_05,
       classification_col = "clusters_res05",
       assay = "SpaNorm")
```

#### Cluster composition (resolution 0.5)

Cell-type calls per cluster, read off the top-10 FindAllMarkers entries above; ordered by lineage rather than by cluster id.

| Lineage | Cluster | Cell type | Top markers |
|---|---|---|---|
| Oligodendrocyte | 0 | Mature OL, high-FC | Mag / Aspa / Mog / Gatm / Ermn / Cldn11 |
| Oligodendrocyte | 9 | Mature OL, lower-FC twin (Slide1-restricted) | Mag / Aspa / Mog / Gatm / Ermn / Cldn11 (FCs 2.6-3.3) |
| Oligodendrocyte | 12 | OPCs | Pdgfra / Tnr / Plpp4 / Gpt2 / Chst11 / Dpyd |
| Astrocyte | 4 | Gray-matter Aqp4+/Gja1+ | Plpp3 / Htra1 / Gja1 / Prodh / Aqp4 / Slc7a10 |
| Astrocyte | 8 | Agt+ subtype | Agt / Slc7a10 / Gja1 / Aqp4 / Pla2g7 |
| Microglia | 2 | Mixed homeostatic + DAM | Cx3cr1 / Gpr34 / Hexb / P2ry12 / Tmem119 / Siglech (Cst7 / Cd14 / Trem2 in tail) |
| Vascular | 5 | Endothelial cells (BBB) | Cldn5 / Cxcl12 / Flt1 / Emcn / Clec2d / Klf4 |
| Vascular | 7 | VLMC + perivascular macrophages | Tagln / Slc47a1 / Mrc1 / Lum / Col1a2 / Mgp |
| Vascular | 15 | Pericytes / mural | Kcnj8 / Atp13a5 / Vtn / Bgn |
| Ventricular | 17 | Ependymal | Ccdc153 / Tmem212 / Dnah11 |
| Ventricular | 18 | Choroid plexus + MHC-II | Car12 / Ltc4s / Adh1 / Tgfbi / H2-Eb1 |
| Excitatory | 1 | Deep-layer Fezf2+ | Fezf2 / Slc17a7 / Syt6 / Oprk1 / Egr1 |
| Excitatory | 3 | Calb2+/Nts+/Tac1+ mixed | Gck / Nts / Calb2 / Tac1 / Slc17a6 |
| Excitatory | 6 | Cux2+/Lamp5+ upper-layer (IEG-rich) | Lamp5 / Pparg / Rspo1 / Tcap / Egr1 / Calb1 / Cux2 |
| Excitatory | 10 | Vglut2+/Rorb+ thalamic relay | Slc17a6 / Plcb4 / Calb2 / Rorb |
| Excitatory | 19 | Tshz2+ deep / hippocampal | Tshz2 / Dpp4 / Nts / Rspo1 |
| Excitatory | 21 | Wif1+/Syt6+ | Wif1 / Syt6 / Calb2 / Sncg |
| Excitatory | 13 | Pou3f1+ stressed neurons | Pou3f1 / Grin2a / Chst1 / Grin2b / Opcml |
| GABAergic | 11 | Sst+/Lhx6+ MGE-derived | Lhx6 / Sst / Reln / Gad1 / Ndnf |
| GABAergic | 14 | Striatal-flavored Tac1+/Adcy5+/Pax6+ | Ido1 / Otof / Tac1 / Adcy5 / Pax6 |
| GABAergic | 20 | Vip+/Crh+ CGE-derived | Vip / Crh / Pthlh / Cnr1 / Gad1 |
| GABAergic | 22 | Pvalb+/Trh+ | Trh / Pvalb / Gad1 / Spp1 |
| Small / atypical | 16 | Lypd1+/Trh+/Slc17a7+ neuron-like | Lypd1 / Trh / Otof / Tshz2 / Slc17a7 |
| Small / atypical | 23 | Multi-lineage doublets (Slide2-private) | Slc32a1 / Mbp / Cldn11 (FCs ≤ 1.1) |
| Immune | 24 | T cells / cytotoxic lymphocytes | Ccl5 / Cd3d / Itk / Nkg7 / Ms4a4b / Gzmb |

## Louvain at resolution 0.7

The 0.7 labels live in `xen$clusters_res07` and were written by the build chunk; we re-color the plots and re-derive markers from that column without rerunning `FindClusters`.

```{r}
#| label: palette-07
# Re-size the palette to the cluster count produced at resolution 0.7.
colorines_07 <- scCustomize_Palette(num_groups = n_distinct(xen$clusters_res07), ggplot_default_colors = FALSE)
```

### UMAP

```{r}
#| label: umap-07
# Same UMAP coordinates as the 0.5 panel; only the cluster colors / boundaries change.
DimPlot(xen, group.by = "clusters_res07", cols = colorines_07, label = TRUE, label.size = 5) + NoLegend()
```

**Result.** 27 clusters at Louvain 0.7, two more than the 25 at 0.5. The full cell-type table sits after the violin panel below. The differences vs the 0.5 cut:

- **Microglia split into homeostatic (cluster 11: Tmem119 / P2ry12 / Selplg / Siglech / Il1a / Cx3cr1) and DAM (cluster 5: Cst7 / Ccl3 / Clec7a / Itgax / Ch25h / Lyz2 / Trem2)**. At 0.5 these share one cluster (2). This split is the load-bearing 0.7 gain for the E4-vs-E2 contrast.
- **A third oligodendrocyte state separates** as cluster 20 (Aspa / Mag / Mog / Gatm / Pllp / Cldn11 at FCs 1.5-1.7), on top of cluster 0 (mature high-FC) and cluster 7 (the Slide1-restricted lower-FC twin).
- The Slide2-private multi-lineage doublet cluster (res05 23) is absorbed into other clusters and disappears as a discrete island.
- The T-cell cluster reproduces at 0.7 (cluster 26) with the same identity and absolute counts as at 0.5.

All other identities map one-to-one between the two resolutions; see the cluster-composition table after the violin panel for the mapping.

### Per slide

```{r}
#| label: umap-07-slide
#| fig-width: 10
#| column: page

# Slide split at the higher resolution; re-check that no recovered sub-cluster is slide-specific.
DimPlot(xen, group.by = "clusters_res07", split.by = "slide", cols = colorines_07) + NoLegend()
```

**Result.** The slide pattern at 0.7 matches the 0.5 pattern almost exactly because the slide-imbalanced clusters survive the higher-resolution split intact:

- **Cluster 7 (Slide1 OL lower-FC twin) is Slide1-private:** 19,758 cells on Slide1 vs 1,194 on Slide2 (ratio 0.06:1, ~33x off the 2:1 baseline). Same identity as cluster 9 at 0.5; Louvain at 0.7 did not redistribute these cells.
- **Cluster 0 (mature OL high-FC) is Slide2-enriched:** 3,409 on Slide1 vs 42,113 on Slide2 (ratio 12.35:1, ~6x over baseline). Stronger Slide2 dominance than at 0.5 because the third OL state (cluster 20 here) carved 5,337 balanced cells out of the high-FC pool.
- The DAM cluster (5) is Slide2-enriched at 4.78:1 (3,657 vs 17,486), ~2.4x over baseline. I read this as slide-correlated DAM load: Slide2 captures more amyloid-proximal microglia in absolute numbers, not as a slide-effect on the DAM signature itself.
- The Pou3f1+ stressed-neuron cluster (15) is Slide2-enriched at 4.68:1 (2,193 vs 10,264), driven by one sample (see below).
- The T-cell cluster (26) is Slide2-enriched at 4.68:1 (94 vs 440); same ratio and same absolute counts as at 0.5. The cluster reproduces on both slides.
- Every other cluster sits within a 1.0-2.9x window of baseline. No 0.7 split introduces a new slide-private cluster.

The Slide2-private doublet cluster (res05 cluster 23) is absorbed at 0.7 and stops being a discrete island; the 0.7 cut gets rid of the smallest slide-private fragment from the 0.5 build for free.

### Per sample

```{r}
#| label: umap-07-sample
#| fig-width: 14
#| fig-height: 12
#| column: page

# Per-sample facets; 4s2M_F2 still expected to be sparse.
DimPlot(xen, group.by = "clusters_res07", split.by = "sample_id", ncol = 3, cols = colorines_07) + NoLegend()
```

**Result.** The per-sample pattern at 0.7 reproduces the 0.5 pattern.  

* Cluster 7 (Slide1 OL twin) lives in 4s2_F1 (7,436), 4s2M_F1 (7,566) and 4s2M_F2 (4,756) and is near-absent on the four Slide2 samples (≤ 599 cells each).  
* Cluster 15 (Pou3f1+ stressed neurons) is dominated by 4s2M_F3 (4,274) and stays under 2,225 cells in every other complete section. 4s2M_F2 is sparser across all clusters as expected for the partial section.  
* The two new biological splits (microglia DAM-vs-homeostatic clusters 5 and 11; third OL state cluster 20) appear in every complete sample, so neither split is driven by one animal.

### Find all markers

```{r}
#| label: find-all-markers-07
#| eval: false

# presto-backed Wilcoxon; identical setup to the 0.5 chunk above.
library(presto)
Idents(xen) <- "clusters_res07"
spanorm_markers_07 <- FindAllMarkers(xen, only.pos = TRUE, densify = TRUE)

write_csv(spanorm_markers_07, file = "results/20260516-find_all_markers_spanorm_p025_clean_res07.csv")
```

```{r}
#| label: markers-table-07

# Reload the cached resolution-0.7 marker table.
spanorm_markers_07 <- read_csv("results/20260516-find_all_markers_spanorm_p025_clean_res07.csv")

spanorm_markers_07 <- spanorm_markers_07 |>
  select(cluster, gene, avg_log2FC, p_val_adj) |>
  group_by(cluster) |>
  arrange(desc(avg_log2FC), .by_group = TRUE)

kbl(spanorm_markers_07 |> slice_head(n = 10)) |>
  kable_styling("striped") |>
  scroll_box(height = "500px")
```

```{r}
#| label: markers-violin-07
#| column: page
#| fig-height: 25
#| fig-width: 12

# Top 5 markers per cluster at resolution 0.7 feed fiddle().
top_5_genes_07 <- spanorm_markers_07 |>
  slice_head(n = 5) |>
  pull(gene) |>
  unique()

fiddle(xen,
       genes_of_interest = top_5_genes_07,
       classification_col = "clusters_res07",
       assay = "SpaNorm")
```

#### Cluster composition (resolution 0.7)

Cell-type calls per cluster, read off the top-10 FindAllMarkers entries above; ordered by lineage. The right-most column carries the matched cluster id at resolution 0.5 for cross-reference.

| Lineage | Cluster | Cell type | Top markers | ↔ res 0.5 |
|---|---|---|---|---|
| Oligodendrocyte | 0 | Mature OL, high-FC | Mag / Aspa / Gatm / Mog / Ermn / Cldn11 | 0 |
| Oligodendrocyte | 7 | Mature OL, lower-FC twin (Slide1-restricted) | Cldn11 / Ermn / Apod / Mog / Aspa / Plin4 | 9 |
| Oligodendrocyte | 20 | Third OL state, low-FC (NEW at 0.7) | Aspa / Mag / Mog / Gatm / Pllp / Cldn11 (FCs 1.5-1.8) | (merged into 9 at 0.5) |
| Oligodendrocyte | 14 | OPCs | Pdgfra / Tnr / Plpp4 / Gpt2 / Chst11 | 12 |
| Astrocyte | 2 | Gray-matter Aqp4+/Gja1+ | Plpp3 / Htra1 / Gja1 / Prodh / Aqp4 / Slc7a10 | 4 |
| Astrocyte | 8 | Agt+ subtype | Agt / Slc7a10 / Gja1 / Aqp4 / Pla2g7 | 8 |
| Microglia | 5 | **DAM microglia (NEW at 0.7)** | Cst7 / Ccl3 / Clec7a / Itgax / Ch25h / Lyz2 / Trem2 | (merged with 11 into 2 at 0.5) |
| Microglia | 11 | **Homeostatic microglia (NEW at 0.7)** | Tmem119 / P2ry12 / Selplg / Siglech / Il1a / Cx3cr1 | (merged with 5 into 2 at 0.5) |
| Vascular | 4 | Endothelial cells (BBB) | Cldn5 / Cxcl12 / Flt1 / Emcn / Clec2d / Klf4 | 5 |
| Vascular | 6 | VLMC + perivascular macrophages | Tagln / Slc47a1 / Mrc1 / Lum / Col1a2 / Mgp | 7 |
| Vascular | 18 | Pericytes / mural | Kcnj8 / Atp13a5 / Vtn / Bgn | 15 |
| Ventricular | 19 | Ependymal | Ccdc153 / Tmem212 / Dnah11 | 17 |
| Ventricular | 21 | Choroid plexus + MHC-II | Car12 / Ltc4s / Adh1 / Tgfbi / H2-Eb1 | 18 |
| Excitatory | 1 | Deep-layer Fezf2+ | Fezf2 / Syt6 / Slc17a7 / Oprk1 / Egr1 | 1 |
| Excitatory | 3 | Calb2+/Nts+/Tac1+ mixed | Gck / Nts / Calb2 / Tac1 / Slc17a6 | 3 |
| Excitatory | 9 | Cux2+/Lamp5+ upper-layer | Lamp5 / Pparg / Stard8 / Otof / Calb1 / Cux2 | 6 (partial) |
| Excitatory | 17 | Rspo1+/Tcap+/IEG-rich | Rspo1 / Tcap / Egr1 / Rorb / Whrn | 6 (partial) |
| Excitatory | 10 | Vglut2+/Rorb+ thalamic relay | Slc17a6 / Plcb4 / Calb2 / Rorb | 10 |
| Excitatory | 22 | Tshz2+ deep / hippocampal | Tshz2 / Dpp4 / Rspo1 / Pparg / Htr2c | 19 |
| Excitatory | 24 | Wif1+/Syt6+ | Wif1 / Syt6 / Calb2 / Sncg / Tac1 | 21 |
| Excitatory | 15 | Pou3f1+ stressed neurons | Pou3f1 / Grin2a / Chst1 / Grin2b / Opcml | 13 |
| GABAergic | 13 | Sst+/Lhx6+ MGE-derived | Lhx6 / Sst / Reln / Gad1 / Ndnf | 11 |
| GABAergic | 16 | Striatal-flavored Tac1+/Adcy5+ | Ido1 / Otof / Tac1 / Adcy5 / Cd36 | 14 |
| GABAergic | 23 | Vip+/Crh+ CGE-derived | Vip / Crh / Pthlh / Cnr1 / Gad1 | 20 |
| GABAergic | 25 | Pvalb+/Trh+ | Trh / Pvalb / Gad1 / Spp1 | 22 |
| Small / atypical | 12 | Lypd1+/Slc17a7+ neuron-like | Lypd1 / Slc17a7 / Reln / Trh / Plpp4 | 16 |
| Immune | 26 | T cells / cytotoxic lymphocytes | Ccl5 / Cd3d / Itk / Nkg7 / Ms4a4b / Gzmb | 24 |





## References

- Salim A et al. "SpaNorm: spatially-aware normalization for spatial transcriptomics data." Genome Biology, 2025. DOI: 10.1186/s13059-025-03565-y.
- SpaNorm vignette: https://bhuvad.github.io/SpaNorm/articles/SpaNorm.html.
- Allen Brain Atlas, mouse: https://mouse.brain-map.org.
- Parent notebook: `20260513_SpaNorm_samplep025.qmd`.
