---
title: "SpaNorm clean: sample.p = 0.25"
subtitle: " cluster 20 + 25 removed"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
---

## Why this notebook

The promoted SpaNorm build at `sample.p = 0.25` (`20260513_SpaNorm_samplep025.qmd`) yields 25 clusters at Louvain 0.5. Two of them, `seurat_clusters` 20 and 25, look like slide-specific debris on the per-slide UMAP and carry top markers we read as artefactual rather than biological. We re-fit per-sample SpaNorm at `sample.p = 0.25` on the raw merged object after dropping the cells assigned to those two clusters, and we re-cluster from PCA forward. 

Every other parameter is held fixed against the parent notebook so the before-after comparison is clean: same seven per-sample SpaNorm fits, same `sample.p = 0.25`, same `df.tps = 6`, same `ScaleData -> PCA(30) -> FindNeighbors -> Louvain(0.5) -> UMAP` chain, same 480-gene variable-feature set.

Bibliography for this notebook:

- Salim et al., "SpaNorm: spatially-aware normalization for spatial transcriptomics data," Genome Biology, 2025. DOI: 10.1186/s13059-025-03565-y.
- SpaNorm package vignette: https://bhuvad.github.io/SpaNorm/articles/SpaNorm.html.
- Parent notebook: `20260513_SpaNorm_samplep025.qmd` (promoted project default).

## Setup

```{r}
#| label: setup

# Seurat v5 holds the merged Xenium object with the per-slide FOV slots we read centroids from.
library(Seurat)
library(tidyverse)
# qs2 reads the multi-GB checkpoint roughly 10x faster than saveRDS.
library(qs2)
library(fs)
# SpaNorm is the spatially aware normalization (Salim, Genome Biology, 2025).
library(SpaNorm)
# SpatialExperiment is the native input class for SpaNorm; we build one per sample.
library(SpatialExperiment)
# scCustomize only contributes its categorical palette helper for the UMAP plots.
library(scCustomize)
# kableExtra renders the per-cluster top-marker table inside a scroll box.
library(kableExtra)

# Source the shared helper functions; fiddle() draws the per-cluster violin panel.
source("R/20260514-helper_functions.R")

# Seed Louvain and any other stochastic step so cluster labels stay stable across renders.
set.seed(1)

# Reload the post-SpaNorm clean checkpoint; built by the chunk below on the one-pass eval:true render.
xen <- qs_read("data/20260516_microglia_switch_spanorm_p025_clean.qs2")
```

## Build the clean SpaNorm assay (drop clusters 20 + 25, re-fit per sample)

```{r}
#| label: spanorm-clean
#| eval: false


# Sample, slide, and condition lookup; matches R/20260512_xenium2seurat.R for self-containment.
samples <- tribble(
  ~sample_id, ~slide,            ~condition,
  "4s2_F1",   "Slide1_0022474",  "control",
  "4s2M_F1",  "Slide1_0022474",  "switched",
  "4s2M_F2",  "Slide1_0022474",  "switched",
  "4s2_F2",   "Slide2_0069080",  "control",
  "4s2_F3",   "Slide2_0069080",  "control",
  "4s2M_F3",  "Slide2_0069080",  "switched",
  "4s2M_F4",  "Slide2_0069080",  "switched")

# Start from the promoted dirty build and drop the rubbish clusters in one go.
# SpaNorm needs raw counts, so we strip the object back to the Xenium assay and drop the
# stale SpaNorm normalization, PCA / UMAP, and seurat_clusters before re-fitting.
xen <- qs_read("data/20260512_microglia_switch_spanorm_p025.qs2")
DefaultAssay(xen) <- "Xenium"
xen <- subset(xen, subset = seurat_clusters %in% c("20", "25"), invert = TRUE)
xen <- DietSeurat(xen, assays = "Xenium", dimreducs = NULL, graphs = NULL)

# Slide -> FOV is fixed: the merge of the two LoadXenium slides keeps the FOVs as fov and fov.2.
slide_to_fov <- c(Slide1_0022474 = "fov", Slide2_0069080 = "fov.2")
fov_lookup   <- set_names(slide_to_fov[samples$slide], samples$sample_id)

# Run SpaNorm on one sample at the package-default subsample. Returns the log-scale normalized matrix.
spanorm_one_sample <- function(s, xen, fov_lookup) {
  # Subset keeps only the cells of one tissue section; SpaNorm's smoothness assumption is per section.
  seu_s <- subset(xen, subset = sample_id == s)
  # Centroids come from the slide FOV; we filter to the cells that survived the per-sample subset.
  coords <- GetTissueCoordinates(seu_s, image = fov_lookup[[s]]) |>
    filter(cell %in% Cells(seu_s))
  # SpatialExperiment is SpaNorm's native input; we hand it raw counts plus the 2D centroids.
  spe <- SpatialExperiment(
    assays        = list(counts = GetAssayData(seu_s, assay = "Xenium", layer = "counts")),
    spatialCoords = as.matrix(coords[, c("x", "y")]))
  # sample.p = 0.25 and df.tps = 6 are held identical to the parent notebook for a like-for-like comparison.
  spe <- SpaNorm(spe,
                 sample.p   = 0.25,
                 df.tps     = 6,
                 adj.method = "auto",
                 backend    = "auto",
                 verbose    = TRUE)
  # logcounts holds the normalized matrix; we keep Seurat's cell barcodes for the later cbind.
  out <- SummarizedExperiment::assay(spe, "logcounts")
  colnames(out) <- Cells(seu_s)
  out}

# Loop over the 7 samples; map keeps the per-sample matrices named for the reduce step.
norm_list <- map(samples$sample_id, spanorm_one_sample,
                 xen = xen, fov_lookup = fov_lookup) |>
  set_names(samples$sample_id)

# qs_save(norm_list, "data/20260516_norm_list_sample025_clean.qs2")

# Bind the per-sample matrices into one wide normalized matrix and reorder to the merged object's cell order.
# Qualify reduce() because SpatialExperiment loads IRanges, whose reduce() masks purrr::reduce.
norm_mat <- purrr::reduce(norm_list, cbind)
norm_mat <- norm_mat[, Cells(xen)]

# Attach as a new v5 assay; data layer only because the auto-selected adjustment returns log-scale residuals.
# CreateAssay5Object avoids the "counts layer not present" warning that CreateAssayObject (v3) triggers inside a v5 object.
xen[["SpaNorm"]] <- CreateAssay5Object(data = norm_mat)
DefaultAssay(xen) <- "SpaNorm"

# Use all 480 genes as variable features; on a targeted Xenium panel every gene was chosen as informative by design,
# and FindVariableFeatures("vst") requires a counts layer the SpaNorm assay does not carry.
VariableFeatures(xen) <- rownames(xen)

# RunPCA expects a scale.data layer; SCT writes it implicitly, but our custom SpaNorm assay only carries the data layer.
xen <- ScaleData(xen, verbose = FALSE)

# Default npcs (50) is fine; we only use the first 30 downstream.
xen <- RunPCA(xen, npcs = 30, verbose = FALSE)

# 1:30 PCs and resolution 0.5 match the parent notebook for a like-for-like comparison.
xen <- FindNeighbors(xen, reduction = "pca", dims = 1:30, verbose = FALSE)
xen <- FindClusters(xen, resolution = 0.5, algorithm = 1, verbose = FALSE)
xen <- RunUMAP(xen, reduction = "pca", dims = 1:30, verbose = FALSE)

qs_save(xen, "data/20260516_microglia_switch_spanorm_p025_clean.qs2")
```

## PCA split by slide

```{r}
#| label: palette
# Palette sized to the actual cluster count so colors stay stable if removing clusters 20+25 shifts the count.
colorines <- scCustomize_Palette(num_groups = n_distinct(xen$seurat_clusters), ggplot_default_colors = FALSE)
```

```{r}
#| label: pca
# Split by slide to compare against the dirty PCA in 20260513_SpaNorm_samplep025.html.
PCAPlot(xen, split.by = "slide", cols = colorines)
```

**Result.** The PC1/PC2 envelope keeps the fan shape of the parent build. Both slides span the same PC1 range (-12 to ~15) and the same PC2 range (-7 to ~28), and the red upper-PC2 lobe (cluster 2 here, microglia per the markers below) sits in matched positions on Slide1 and Slide2. The dense low-PC2 multi-cluster basin reproduces on both panels with no slide-only outlier. The cluster-20 / cluster-25 cells removed before re-fitting do not leave a visible hole in the PCA envelope.

## UMAP

```{r}
#| label: umap
# label = TRUE prints the cluster id inside each UMAP island so the markers panels below stay legible.
DimPlot(xen, cols = colorines, label = TRUE, label.size = 5) + NoLegend()
```

**Result.** Cleaning yields 22 clusters at Louvain 0.5, three fewer than the 25-cluster parent build. The drop is one cluster more than the two we removed, which tells us the cluster-20 / cluster-25 cells were not standalone islands; their removal redistributed cells across neighbours and Louvain collapsed one further cluster at the same resolution. We therefore treat the new decomposition as a fresh cluster set, not a trimmed version of the 25-cluster map, and re-run FindAllMarkers below to re-derive cell-type calls.

The 22 islands span the same UMAP footprint as the parent (umap_1 from -15 to ~10, umap_2 from -10 to ~13). All major basins reappear: the oligodendrocyte mass at the bottom (cluster 0), the microglia lobe (cluster 2 here), the astrocyte cluster 4, the endothelial cluster 7, and the satellite islands at the periphery (clusters 16, 18, 19) all reproduce the parent topology.

### Per slide

```{r}
#| label: umap-slide
#| fig-width: 10
#| column: page

# Side-by-side UMAPs per slide; the failure mode we test against is any residual cluster that lives on only one slide.
DimPlot(xen, split.by = "slide", cols = colorines) + NoLegend()
```

**Result.** The two slides now populate the same cluster geometry across the whole UMAP. Every island that carried cells on the parent build still does so on both slides, with comparable densities in the large basins (oligodendrocytes at the bottom, microglia mid-left, astrocytes centre, excitatory neurons upper-right). The slide imbalance the parent build still showed at the bottom-centre lobe is no longer visible after the cluster-20 / cluster-25 removal. The one residual asymmetry is the top-centre satellite (cluster 16, ciliated/ependymal per Ccdc153 / Tmem212 / Dnah11 in the violin panel below), which sits clearly denser on Slide2 than on Slide1; the markers are anatomically plausible (ependyma lines the ventricles, which sample differently between the two slides), so we treat this as a sampling effect rather than a normalization residual and leave it in place.

### Per sample

```{r}
#| label: umap-sample
#| fig-width: 14
#| fig-height: 12
#| column: page

# Per-sample facets; 4s2M_F2 (partial section) is still expected to contribute fewer cells than the other six.
DimPlot(xen, split.by = "sample_id", ncol = 3, cols = colorines) + NoLegend()
```

**Result.** All six complete sections (4s2_F1, F2, F3 and 4s2M_F1, F3, F4) reproduce the 22-cluster landscape with similar proportions. 4s2M_F2 contributes fewer cells, as expected for the partial section, but its cells land in the same basins as the matched complete sections with no new sample-only island. The cluster-16 satellite (ependymal) shows the same Slide1-versus-Slide2 split flagged above: it is absent from 4s2_F1, 4s2M_F1, and 4s2M_F2 (all Slide1) and clearly populated in 4s2_F2, 4s2_F3, 4s2M_F3, and 4s2M_F4 (all Slide2), consistent with a tissue-sampling effect on ventricle-adjacent ependymal cells rather than a normalization artefact.

## Find all markers

```{r}
#| label: find-all-markers
#| eval: false

# devtools::install_github('immunogenomics/presto')
# presto gives a far faster Wilcoxon Rank Sum implementation than Seurat's default.
library(presto)

# FindAllMarkers needs cluster identities as the active Idents().
Idents(xen) <- "seurat_clusters"
spanorm_markers <- FindAllMarkers(xen, only.pos = TRUE, densify = TRUE)

 write_csv(spanorm_markers, file = "results/20260516-find_all_markers_spanorm_p025_clean_res05.csv")
```


```{r}
#| label: markers-table

# Reload the cached marker table so the table chunk renders without rerunning FindAllMarkers.
spanorm_markers <- read_csv("results/20260516-find_all_markers_spanorm_p025_clean_res05.csv")

# Top 10 markers per cluster, sorted by avg_log2FC inside each cluster.
spanorm_markers <- spanorm_markers |>
  select(cluster, gene, avg_log2FC, p_val_adj) |>
  group_by(cluster) |>
  arrange(desc(avg_log2FC), .by_group = TRUE)


kbl(spanorm_markers |> slice_head(n = 10)) |>
  kable_styling("striped") |>
  scroll_box(height = "500px")

```

```{r}
#| label: markers-violin
#| column: page
#| fig-height: 25
#| fig-width: 12

# Top 5 markers per cluster feed fiddle(); the SpaNorm assay carries the log-scale normalized values.
top_5_genes_spanorm <- spanorm_markers |>
  slice_head(n = 5) |>
  pull(gene) |>
  unique()

fiddle(xen,
       genes_of_interest = top_5_genes_spanorm,
       classification_col = "seurat_clusters",
       assay = "SpaNorm")
```


## Reclustering at higher Louvain resolution (0.7)

The cell-type assignment table for the 22-cluster, resolution-0.5 decomposition (`notes/20260516-cluster_celltype_assignments_spanorm_clean.md`) flags two regressions against the parent build: the discrete cytotoxic-T-cell cluster (parent 24: Cd3d / Gzmb / Nkg7 / Itk) is gone, and the parent's Sst+, Vip+, and Lamp5+/Lypd1+ inhibitory subclasses collapse into one mixed GABAergic cluster. Both look like under-clustering rather than a true loss of signal: lymphocyte markers redistribute into the VLMC / perivascular cluster at pct.1 ≤ 0.03, and the merged inhibitory cluster carries Vip + Sst + Lhx6 + Pvalb together in its top-10. We re-run `FindClusters` at resolution 0.7 on the same clean PCA / SNN graph to test whether the higher resolution recovers the T cells and re-splits the inhibitory subclasses without disturbing the rest of the landscape.

```{r}
#| label: cluster-07



# Same SNN graph, higher Louvain resolution; this overwrites seurat_clusters with the 0.7 labels.
# The 0.5 labels remain accessible in xen$SpaNorm_snn_res.0.5 if needed for a side-by-side check.
xen <- FindClusters(xen, resolution = 0.7, algorithm = 1, verbose = FALSE)


```

```{r}
#| label: palette-07
# Re-size the palette to the new cluster count produced at resolution 0.7.
colorines_07 <- scCustomize_Palette(num_groups = n_distinct(xen$seurat_clusters), ggplot_default_colors = FALSE)
```

### UMAP

```{r}
#| label: umap-07
# Same UMAP coordinates as the 0.5 panel; only the cluster colors / boundaries change.
DimPlot(xen, cols = colorines_07, label = TRUE, label.size = 5) + NoLegend()
```

**Result.** Resolution 0.7 produces 30 clusters, up from 22 at resolution 0.5. The two regressions flagged in the cell-type assignment table are recovered:

1. **T cells reappear as a discrete cluster (28).** Cd3d at FC 7.52 (pct.1 0.72), Ccl5 at 7.37, Itk at 6.57, Nkg7 at 6.15, Ms4a4b at 5.73, Tnf at 4.94, Gzmb at 4.93. The signal is at least as strong as in the parent SpaNorm cluster 24 and sits as a discrete small island at the right edge of the UMAP.
2. **GABAergic subclasses re-split.** Cluster 17 carries Lhx6 (5.81) + Sst (5.46) + Gad1 (3.44), cluster 21 carries Vip (8.50) + Crh (6.08) + Ndnf (4.60) + Reln (4.26), and cluster 26 carries Pvalb (5.71) + Trh (6.64) + Gad1 (4.40). The mixed inhibitory cluster from 0.5 is gone.

Two further gains come from the higher resolution that the 0.5 build also missed:

3. **Microglia split into homeostatic and DAM.** Cluster 12 is homeostatic microglia (Tmem119 5.43, P2ry12 5.20, Selplg 4.62, Siglech 4.58, Cx3cr1 4.50). Cluster 6 is DAM (Cst7 6.37, Ccl3 6.02, Itgax 5.57, Clec7a 5.57, Ch25h 5.35, Lyz2 5.28, Bcl2a1b 5.09, Cd14 4.97, Trem2 4.80). This split was not visible in either the parent SpaNorm build or the 0.5 clean build; both held microglia as a single mixed cluster.
4. **Three excitatory neuron subpopulations resolved.** Cluster 24 recovers the parent's Tshz2+ neurons (Tshz2 4.22, Dpp4 3.44). Cluster 25 is a Wif1+/Syt6+/Calb2+/Sncg+ excitatory population (Wif1 5.80, Syt6 5.13). Cluster 29 is a Nts+/Fezf2+/Pou3f1+ deep-layer cluster (Nts 5.89, Fezf2 3.64).

The large basins (oligodendrocyte at cluster 0, mature OL low-FC twin at 9, OPCs at 13, astrocyte at 2 and 8, endothelial at 4, pericyte at 18, VLMC at 7, ependymal at 20, choroid plexus at 22) keep their identity and only fragment internally where biologically warranted (microglia 6/12; new third OL state at 19). The 0.7 decomposition is the working scaffold for the E4-vs-E2 DE.

### Per slide

```{r}
#| label: umap-07-slide
#| fig-width: 10
#| column: page

# Slide split at the new resolution; we re-check that no recovered subcluster is slide-specific.
DimPlot(xen, split.by = "slide", cols = colorines_07) + NoLegend()
```

**Result.** Both slides span the same UMAP footprint and populate every basin at comparable densities. The recovered T-cell island (cluster 28, right edge) appears on both slides. The recovered Vip+ (21), Sst/Lhx6+ (17), and Pvalb/Trh+ (26) inhibitory subclusters all populate Slide1 and Slide2 alike. The only residual asymmetry is the ependymal cluster (20, top-centre satellite), which stays enriched on Slide2 for the same ventricle-sampling reason flagged at resolution 0.5. The higher Louvain resolution does not introduce any new slide-specific cluster.

### Per sample

```{r}
#| label: umap-07-sample
#| fig-width: 14
#| fig-height: 12
#| column: page

# Per-sample facets; 4s2M_F2 still expected to be sparse.
DimPlot(xen, split.by = "sample_id", ncol = 3, cols = colorines_07) + NoLegend()
```

**Result.** All six complete sections (4s2_F1, F2, F3 and 4s2M_F1, F3, F4) reproduce the 30-cluster landscape with comparable proportions. 4s2M_F2 contributes fewer cells, as expected for the partial section, but the cells it does contribute land in the shared basins with no new sample-only island. The recovered T-cell cluster (28) and the three inhibitory subclusters (17 Sst+/Lhx6+, 21 Vip+, 26 Pvalb+) all appear in every complete section, so neither recovery is driven by one sample. The DAM microglia cluster (6) and the homeostatic microglia cluster (12) also populate every complete section, which means the homeostatic-DAM split is not an artefact of one animal's plaque load.

### Find all markers (resolution 0.7)

```{r}
#| label: find-all-markers-07
#| eval: false

# presto-backed Wilcoxon; identical setup to the 0.5 chunk above.
library(presto)
Idents(xen) <- "seurat_clusters"
spanorm_markers_07 <- FindAllMarkers(xen, only.pos = TRUE, densify = TRUE)

write_csv(spanorm_markers_07, file = "results/20260516-find_all_markers_spanorm_p025_clean_res07.csv")
```

```{r}
#| label: markers-table-07

# Reload the cached resolution-0.7 marker table so the table renders without rerunning FindAllMarkers.
spanorm_markers_07 <- read_csv("results/20260516-find_all_markers_spanorm_p025_clean_res07.csv")

# Top 10 markers per cluster at resolution 0.7, sorted by avg_log2FC inside each cluster.
spanorm_markers_07 <- spanorm_markers_07 |>
  select(cluster, gene, avg_log2FC, p_val_adj) |>
  group_by(cluster) |>
  arrange(desc(avg_log2FC), .by_group = TRUE)

kbl(spanorm_markers_07 |> slice_head(n = 10)) |>
  kable_styling("striped") |>
  scroll_box(height = "500px")
```

```{r}
#| label: markers-violin-07
#| column: page
#| fig-height: 25
#| fig-width: 12

# Top 5 markers per cluster at resolution 0.7 feed fiddle().
top_5_genes_spanorm_07 <- spanorm_markers_07 |>
  slice_head(n = 5) |>
  pull(gene) |>
  unique()

fiddle(xen,
       genes_of_interest = top_5_genes_spanorm_07,
       classification_col = "seurat_clusters",
       assay = "SpaNorm")
```


## Summary

Re-fitting per-sample SpaNorm at `sample.p = 0.25` after dropping the cluster-20 cells from the promoted build yields 22 clusters at Louvain 0.5 and 30 clusters at Louvain 0.7. Per-slide balance improves at both resolutions; the only residual asymmetry is the ciliated/ependymal cluster (Ccdc153 / Tmem212 / Dnah11) enriched on Slide2, which we attribute to ventricle sampling rather than a slide effect on normalization. The 0.5 decomposition preserves the biological backbone (oligodendrocytes, astrocytes, microglia, endothelial, neurons) but loses the discrete cytotoxic-T-cell cluster from the parent build (parent 24) and collapses the parent's Sst+/Vip+/Lamp5+ inhibitory subclasses into one mixed GABAergic cluster. Raising Louvain to 0.7 on the same clean PCA / SNN graph fully recovers the T cells as cluster 28 (Cd3d FC 7.52, Ccl5 7.37, Itk 6.57, Nkg7 6.15, Tnf 4.94, Gzmb 4.93), re-splits the inhibitory subclasses into Sst+/Lhx6+ (17), Vip+/Crh+ (21), and Pvalb+/Trh+ (26), and as a bonus splits microglia into homeostatic (12: Tmem119 / P2ry12 / Selplg / Siglech / Cx3cr1) and DAM (6: Cst7 / Ccl3 / Itgax / Clec7a / Ch25h / Trem2 / Cd14). The DAM split is a new finding that neither the parent SpaNorm build nor the 0.5 clean build resolved. We promote the 0.7 clean decomposition as the working SpaNorm scaffold and re-run the cell-type assignment table against the 30-cluster set before starting the E4-vs-E2 DE.

## References

- Salim A et al. "SpaNorm: spatially-aware normalization for spatial transcriptomics data." Genome Biology, 2025. DOI: 10.1186/s13059-025-03565-y.
- SpaNorm vignette: https://bhuvad.github.io/SpaNorm/articles/SpaNorm.html.
- Parent notebook: `20260513_SpaNorm_samplep025.qmd`.

