---
title: "SpaNorm: spatial-aware normalization"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
---

## Background

The two SCT notebooks already in this repo expose a slide effect on the merged Seurat object (see `20260512_QC.qmd`). `SCTransform(vars.to.regress = "slide")` removes that effect as a single global mean shift. It cannot remove the spatial gradient of library size that runs across each tissue section, and that gradient is the artefact most likely to drive false spatially variable genes in Xenium data (Salim, Genome Biology, 2025).

SpaNorm models the gradient directly. We fit a negative-binomial GLM in which the gene- and location-specific library size is a 2D thin-plate spline of the cell centroids, regress that smooth component out, and write the log-scale adjusted counts back to Seurat as a new `SpaNorm` assay (Salim, Genome Biology, 2025). The model assumes spatial smoothness, so we fit one model per tissue section (seven fits across two slides) rather than one fit per slide.

The installed package version is SpaNorm 1.4.0 (Bioc 3.22), which only dispatches on `SpatialExperiment`; `methods("SpaNorm")` confirms a single `SpaNorm,SpatialExperiment-method` and no Seurat method. The development build on GitHub (`bhuvad/SpaNorm`, 1.5.x) adds a Seurat method that reads centroids from `meta.data$x` / `$y`, but we stay on the Bioconductor release. Therefore the build chunk wraps each per-sample subset in a `SpatialExperiment` before calling `SpaNorm()`, then writes the returned `logcounts` matrix back into a new Seurat assay.

Bibliography for this notebook:

- Salim et al., "SpaNorm: spatially-aware normalization for spatial transcriptomics data," Genome Biology, 2025. DOI: 10.1186/s13059-025-03565-y.
- SpaNorm package vignette: https://bhuvad.github.io/SpaNorm/articles/SpaNorm.html.
- Cached NotebookLM read-through: `notebookLM/2026-05-12_spanorm-background.md`.

## Setup

```{r}
#| label: setup

# Seurat v5 holds the merged Xenium object with the per-slide FOV slots we read centroids from.
library(Seurat)
library(tidyverse)
# qs2 reads the ~5 GB checkpoint roughly 10x faster than saveRDS.
library(qs2)
library(fs)
# SpaNorm is the spatially aware normalization (Salim, Genome Biology, 2025).
library(SpaNorm)
# SpatialExperiment is the native input class for SpaNorm; we build one per sample.
library(SpatialExperiment)
# scCustomize only contributes its categorical palette helper for the UMAP plots.
library(scCustomize)

# Seed Louvain and any other stochastic step so cluster labels stay stable across renders.
set.seed(1)

# Reload the post-SpaNorm checkpoint so renders skip the ~40 min build chunk below.
xen <- qs_read("data/20260512_microglia_switch_spanorm.qs2")
```

## Build the SpaNorm assay (one fit per sample)

```{r}
#| label: spanorm
#| eval: false
# eval:false because this is the one-off build; every render reloads the checkpoint above.

# Sample, slide, and condition lookup; matches R/20260512_xenium2seurat.R for self-containment.
samples <- tribble(
  ~sample_id, ~slide,            ~condition,
  "4s2_F1",   "Slide1_0022474",  "control",
  "4s2M_F1",  "Slide1_0022474",  "switched",
  "4s2M_F2",  "Slide1_0022474",  "switched",
  "4s2_F2",   "Slide2_0069080",  "control",
  "4s2_F3",   "Slide2_0069080",  "control",
  "4s2M_F3",  "Slide2_0069080",  "switched",
  "4s2M_F4",  "Slide2_0069080",  "switched")

# Load the raw merged object; SpaNorm needs raw counts, not the SCT residuals.
xen <- qs_read("data/20260512_microglia_switch_original_seurat.qs2")
DefaultAssay(xen) <- "Xenium"

# Build the sample_id -> FOV-name lookup from the object itself; LoadXenium uses default name "fov" and merge renames collisions
# (so the actual names are usually fov.1 / fov.2, not the slide labels).
fov_to_slide <- map_chr(Images(xen), \(img) {
  # Each FOV's cells share the same slide value in meta.data; use the first one as the slide tag.
  unique(xen@meta.data[Cells(xen[[img]]), "slide"])[1]
}) |> set_names(Images(xen))
# Invert: slide -> FOV, then index by sample slide to get the per-sample FOV lookup.
slide_to_fov <- set_names(names(fov_to_slide), unname(fov_to_slide))
fov_lookup <- set_names(slide_to_fov[samples$slide], samples$sample_id)

# Run SpaNorm on one sample. Returns the log-scale normalized matrix with the original cell barcodes as colnames.
spanorm_one_sample <- function(s, xen, fov_lookup) {
  # Subset keeps only the cells of one tissue section; SpaNorm's smoothness assumption is per section.
  seu_s <- subset(xen, subset = sample_id == s)
  # Centroids come from the slide FOV; we filter to the cells that survived the per-sample subset.
  coords <- GetTissueCoordinates(seu_s, image = fov_lookup[[s]]) |>
    filter(cell %in% Cells(seu_s))
  # SpatialExperiment is SpaNorm's native input; we hand it raw counts plus the 2D centroids.
  spe <- SpatialExperiment(
    assays        = list(counts = GetAssayData(seu_s, assay = "Xenium", layer = "counts")),
    spatialCoords = as.matrix(coords[, c("x", "y")]))
  # sample.p=0.05 follows (Salim, Genome Biology, 2025) for ~400k-cell datasets; auto picks adj.method and backend.
  spe <- SpaNorm(spe,
                 sample.p   = 0.05,
                 df.tps     = 6,
                 adj.method = "auto",
                 backend    = "auto",
                 verbose    = FALSE)
  # logcounts holds the normalized matrix; we keep Seurat's cell barcodes for the later cbind.
  out <- SummarizedExperiment::assay(spe, "logcounts")
  colnames(out) <- Cells(seu_s)
  out}

# Loop over the 7 samples; map keeps the per-sample matrices named for the reduce step.
norm_list <- map(samples$sample_id, spanorm_one_sample,
                 xen = xen, fov_lookup = fov_lookup) |>
  set_names(samples$sample_id)

# Bind the per-sample matrices into one wide normalized matrix and reorder to the merged object's cell order.
# Qualify reduce() because SpatialExperiment loads IRanges, whose reduce() masks purrr::reduce.
norm_mat <- purrr::reduce(norm_list, cbind)
norm_mat <- norm_mat[, Cells(xen)]

# Attach as a new v5 assay; data layer only because the auto-selected adjustment returns log-scale residuals.
# CreateAssay5Object avoids the "counts layer not present" warning that CreateAssayObject (v3) triggers inside a v5 object.
xen[["SpaNorm"]] <- CreateAssay5Object(data = norm_mat)
DefaultAssay(xen) <- "SpaNorm"

# Use all 480 genes as variable features; on a targeted Xenium panel every gene was chosen as informative by design,
# and FindVariableFeatures("vst") requires a counts layer the SpaNorm assay does not carry.
VariableFeatures(xen) <- rownames(xen)

# RunPCA expects a scale.data layer; SCT writes it implicitly, but our custom SpaNorm assay only carries the data layer.
xen <- ScaleData(xen, verbose = FALSE)

# Default npcs (50) is fine; we only use the first 30 downstream.
xen <- RunPCA(xen, npcs = 30, verbose = FALSE)

# 1:30 PCs matches the SCT twins; FindNeighbors default k = 20.
xen <- FindNeighbors(xen, reduction = "pca", dims = 1:30, verbose = FALSE)

# algorithm = 1 is plain Louvain; resolution 0.5 matches the SCT twins for a like-for-like comparison.
xen <- FindClusters(xen, resolution = 0.5, algorithm = 1, verbose = FALSE)

# UMAP only for visual QC of the clusters; not used by any downstream stat.
xen <- RunUMAP(xen, reduction = "pca", dims = 1:30, verbose = FALSE)

# Checkpoint commented out: uncomment to refresh data/20260512_microglia_switch_spanorm.qs2 after parameter changes.
# qs_save(xen, "data/20260512_microglia_switch_spanorm.qs2")
```



## PCA split by slide
```{r}
#| label: palette
# Palette sized to the actual cluster count so colors stay stable if resolution changes the number of clusters.
colorines <- scCustomize_Palette(num_groups = n_distinct(xen$seurat_clusters), ggplot_default_colors = FALSE)
```
```{r}
#| label: pca
# Split by slide to eyeball whether the slide effect documented in 20260512_QC.qmd survives spatial-aware normalization.
PCAPlot(xen, split.by = "slide", cols = colorines)
```

## UMAP

```{r}
#| label: umap
DimPlot(xen, cols = colorines)
```

### Per slide

```{r}
#| label: umap-slide
#| fig-width: 10
#| column: page

# Side-by-side UMAPs per slide reveal any cluster that only exists in one slide; the failure mode we test against.
DimPlot(xen, split.by = "slide", cols = colorines) + NoLegend()
```

### Per sample

```{r}
#| label: umap-sample
#| fig-width: 14
#| fig-height: 12
#| column: page

# Per-sample facets check whether 4s2M_F2 (partial section) lands in the same cluster geometry as the other six.
DimPlot(xen, split.by = "sample_id", ncol = 3, cols = colorines) + NoLegend()
```

## References

- Salim A et al. "SpaNorm: spatially-aware normalization for spatial transcriptomics data." Genome Biology, 2025. DOI: 10.1186/s13059-025-03565-y.
- SpaNorm vignette: https://bhuvad.github.io/SpaNorm/articles/SpaNorm.html.
- Cached NotebookLM read-through: `notebookLM/2026-05-12_spanorm-background.md`.
