---
title: "APOE allele probe validation"
subtitle: ""
date: last-modified
---



* The mBrain_480g Xenium panel carries **four** rs-prefixed probes covering the two SNPs that define the human APOE alleles.   
* All cells express **APOE4/APOE4** in the genomic background.  
* In the 4s2M microglia tamoxifen induces microglia-restricted Cre activation and switches APOE E4/E4 to E2/E2.  
* **The whole pipeline assumes a clean microglia object.** Every result here is computed on the cells inside `data/20260517_microglia_subcluster.qs2`. _If non-microglial cells (cluster/s contamination), they "dilute" the 4s2M switch+ fraction (the Cre-driven cassette is microglia-restricted)_.



---
**APOE allele probes.**

| Probe | Variant at SNP | Present in alleles | Expected in 4s2 (all E4) | Expected in 4s2M microglia (E2) |
|---|---|---|---|---|
| `rs429358-WT-228` | T at rs429358 | E2, E3 | 0 | high (only allele carrying T at rs429358 after switch) |
| `rs429358-ALT-228:C` | C at rs429358 | E4 | high (every cell carries C) | low (switched microglia lost C) |
| `rs7412-WT-226` | C at rs7412 | E3, E4 | high (every cell carries C) | slightly lower (microglia lost C) |
| `rs7412-ALT-226:T` | T at rs7412 | E2 | 0 | high (microglia gained T at rs7412 after switch) |







```{r}
#| label: setup

library(Seurat)
library(tidyverse)
library(qs2)
library(DESeq2)
library(EnhancedVolcano)
library(patchwork)
library(gridExtra)
library(kableExtra)
library(gt)

source("R/20260514-helper_functions.R")
set.seed(1)

mg <- qs_read("data/20260517_microglia_subcluster.qs2")
DefaultAssay(mg) <- "Xenium"
```

## Per-probe counts

```{r}
#| label: probe-pull

# Pull per-cell raw counts of the four probes (both SNPs carry a WT and an ALT variant); reused by
# the histogram chunk below, the summary table that follows, and the ROC fit further down.
probes <- c("rs429358-WT-228", "rs429358-ALT-228:C", "rs7412-WT-226", "rs7412-ALT-226:T")
stopifnot(all(probes %in% rownames(mg[["Xenium"]])))
mm <- as.matrix(GetAssayData(mg, assay = "Xenium", layer = "counts")[probes, ])

# Attach per-cell probe counts to the Seurat object so the ROC fit and the spatial / DE chunks below
# can pull them straight from mg@meta.data without re-extracting the count matrix.
mg$rs429358_wt  <- as.integer(mm["rs429358-WT-228", ])
mg$rs429358_alt <- as.integer(mm["rs429358-ALT-228:C", ])
mg$rs7412_wt    <- as.integer(mm["rs7412-WT-226", ])
mg$rs7412_alt   <- as.integer(mm["rs7412-ALT-226:T", ])
```

### Reads per cell per probe

 

```{r}
#| label: probe-histograms
#| column: page
#| fig-width: 11
#| fig-height: 6

# Long-format count table aligned to per-cell condition; t(mm) puts cells in rows and probes in cols
# so the bound condition vector and the pivot stay straight.
probe_long <- as_tibble(t(mm)) |>
  mutate(condition = mg$condition) |>
  pivot_longer(-condition, names_to = "probe", values_to = "count")

# Log10 y-axis so the 0-read bin stays visible without flattening the small positive tail; bin width 1
# matches the integer count grid. Facet x-scales are free because the four probes have different tails.
probe_long |>
  ggplot(aes(x = count, fill = condition)) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.55) +
  scale_y_log10() +
  scale_fill_manual(values = c(control = "firebrick", switched = "steelblue")) +
  facet_wrap(~ probe, scales = "free_x") +
  labs(x = "Reads per cell", y = "Cell count (log10)",
       title = "Per-cell read distribution for each APOE allele probe",
       subtitle = "Control (4s2, E4/E4 everywhere) vs switched (4s2M, microglia E2/E2)") +
  theme_minimal(base_size = 14)
```

### Proportion (ratio) of reads for each probe


```{r}
#| label: probe-validation

# Per-probe summary: how many cells have >= 1 read, mean count per condition.
# The expected_ctrl column records what we should see under the E4/E4 background.
tibble(
  probe              = probes,
  #snp    = c("T at rs429358 (E2/E3)", "C at rs429358 (E4)",
  #                       "C at rs7412 (E3/E4)", "T at rs7412 (E2)"),
  expected_in_4s2    = c("0", "high (E4 everywhere)",
                         "high (E4 everywhere)", "0"),
  # n_cells_ge1        = rowSums(mm >= 1),
  # n_cells_ge2        = rowSums(mm >= 2),
  # n_cells_ge3        = rowSums(mm >= 3),
  mean_ctrl          = round(rowMeans(mm[, mg$condition == "control"]),  3),
  mean_sw            = round(rowMeans(mm[, mg$condition == "switched"]), 3),
  ratio_sw_ctrl      = round(rowMeans(mm[, mg$condition == "switched"]) /
                             pmax(rowMeans(mm[, mg$condition == "control"]), 1e-3), 2)) |>
  kbl(caption = "Per-probe counts (35,700 cells, microglia subset object).") |>
  kable_styling("striped", full_width = FALSE)
```

**Column legend:**

- `probe`: probe ID exactly as named on the mBrain_480g panel.
- `expected_in_4s2`: prediction under the E4/E4 background, where every cell in the 4s2 control arm is homozygous E4. `0` means the probe should fire in no cell; `high` means it should fire in every cell.
- `mean_ctrl`: mean per-cell count of the probe in the 4s2 control arm.
- `mean_switch`: mean per-cell count of the probe in the 4s2M switched arm.
- `ratio`: `mean_switch / mean_ctrl`. Values above 1 mean the probe rises in the switched arm, values below 1 mean it falls. 




### APOE E2 module score on the microglia UMAP

The Seurat `AddModuleScore()` averages the per-cell expression of the two E2-evidence probes (rs429358-WT-228 + rs7412-ALT-226:T)

```{r}
#| label: module-score
#| column: page
#| fig-width: 12
#| fig-height: 5


# from the same expression bin; the 480-gene panel gives plenty of background controls. We score on
# the per-sample-refit SpaNorm assay (single consolidated data layer) rather than the Xenium counts
# assay (per-sample split layers, which AddModuleScore cannot iterate over in Seurat v5).
# nbin = 10 and ctrl = 30 are tuned for the 480-gene panel; the Seurat defaults (nbin = 24, ctrl = 100)
# overfill each expression bin and trigger a sample-without-replacement error on a panel this small.
mg <- AddModuleScore(mg,
                     features = list(c("rs429358-WT-228", "rs7412-ALT-226:T")),
                     name     = "apoe_E2_module",
                     assay    = "SpaNorm",
                     nbin     = 10,
                     ctrl     = 30)

# AddModuleScore appends a numeric suffix; the metadata column lands as "apoe_E2_module1".
# Split the FeaturePlot by condition so the within-control noise floor and the within-switched
# shift are visible side-by-side. q20 / q95 clipping keeps the colour scale legible against outliers.
FeaturePlot(mg,
            features   = "apoe_E2_module1",
            pt.size= 0.8,
            reduction  = "umap",
            split.by   = "condition",
            cols       = c("grey96", "red"),
            order      = TRUE,
            min.cutoff = "q20",
            max.cutoff = "q98") 
  
```


```{r}
#| label: module-score-sample
#| fig-width: 12
#| fig-height: 15
#| column: page

# FeaturePlot ignores `ncol` when `split.by` is set (it only honours ncol with multiple features).
# Use combine = FALSE to get a list of one panel per sample, then patchwork::wrap_plots(ncol = 2)
# to lay them out on a 2-column grid.
FeaturePlot(mg,
            features   = "apoe_E2_module1",
            pt.size    = 0.8,
            cols       = c("grey96", "red"),
            reduction  = "umap",
            split.by   = "sample_id",
            order      = TRUE,
            min.cutoff = "q20",
            max.cutoff = "q98",
            combine    = FALSE) |>
  patchwork::wrap_plots(ncol = 2)
```



## Per-cell switch call

The two E2-evidence probes (`rs429358-WT-228` and `rs7412-ALT-226:T`)  are combined to create a binary switch+ / switch- call per microglia (**K_raw**).  
The threshold is calibrated on the 4s2 (E4/E4) arm, where every read on either E2 probe shoud be "background noise". We pick the smallest integer cutoff that keeps the 4s2 false-positive (FP) rate at or below 5 %.

### Calibrate the threshold against the 4s2 background

```{r}
#| label: switch-calibration
#| fig-width: 11
#| fig-height: 4

# Per-cell sum of the two T-allele probes; integer-grounded and matches the boss's "express E2" framing.
mg$e2_sum <- mg$rs429358_wt + mg$rs7412_alt

# Empirical FP rate of "e2_sum >= K" on the 4s2 (E4/E4) microglia; pick the smallest K with FP <= 5 %.
e2_sum_ctrl <- mg$e2_sum[mg$condition == "control"]
fp_table <- tibble(K = 1:max(mg$e2_sum)) |>
  mutate(fp_rate = map_dbl(K, \(k) mean(e2_sum_ctrl >= k)))
K_raw <- min(fp_table$K[fp_table$fp_rate <= 0.05])
stopifnot("4s2 background never falls below 5 % at any positive K; lower the FP target." = is.finite(K_raw))

# Parallel calibration on the continuous AddModuleScore output, for cross-checking only.
T_module <- quantile(mg$apoe_E2_module1[mg$condition == "control"], 0.95, na.rm = TRUE)

# Left panel: count histogram of e2_sum, log y so the zero bin and the tail are both visible.
p_count <- mg@meta.data |>
  as_tibble() |>
  ggplot(aes(e2_sum, fill = condition)) +
  geom_histogram(binwidth = 1, position = "identity", alpha = 0.55) +
  scale_y_log10() +
  geom_vline(xintercept = K_raw - 0.5, linetype = "dashed", colour = "black") +
  scale_fill_manual(values = c(control = "firebrick", switched = "steelblue")) +
  labs(x = "rs429358_wt + rs7412_alt\n (raw reads / cell)",
       y = "Cells (log10)",
       title = sprintf("E2-probe count distribution\n (K_raw = %d, FP = %.1f %%)",
                       K_raw, 100 * fp_table$fp_rate[fp_table$K == K_raw])) +
  theme_minimal(base_size = 12)

# Right panel: density of the continuous module score with the 95th-percentile cutoff.
p_mod <- mg@meta.data |>
  as_tibble() |>
  ggplot(aes(apoe_E2_module1, fill = condition)) +
  geom_density(alpha = 0.55, colour = NA) +
  geom_vline(xintercept = T_module, linetype = "dashed", colour = "black") +
  scale_fill_manual(values = c(control = "firebrick", switched = "steelblue")) +
  labs(x = "apoe_E2_module1", y = "Density",
       title = sprintf("Module-score distribution\n (T_module = %.3f)", T_module)) +
  theme_minimal(base_size = 12)

p_count + p_mod + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```


- `K_raw` (in this render, **K_raw = 6**) is the minimum number of E2-probe reads a microglia must carry to be called switch+. We sum the per-cell counts of `rs429358-WT-228` and `rs7412-ALT-226:T`, tally how often 4s2 microglia reach 1, 2, 3, ... reads by chance alone, and pick the smallest cutoff that keeps no more than 5 % of 4s2 microglia at or above it. In our data 3.0 % of 4s2 microglia carry six or more E2 reads by chance, so any 4s2M microglia with six or more reads sits comfortably above the noise floor.

- `T_module` (in this render, **T_module = 0.762**) is the same idea applied to a normalized continuous scorecalculated by `AddModuleScore()`. This function averages the two E2 probes after subtracting, for each cell, the mean expression of 30 randomly chosen panel genes from the same expression bin. The subtraction corrects for cell-to-cell differences in total transcript depth (a sparsely-segmented cell does not look switch+ just because it is small). `T_module` is the 95th percentile of that score on 4s2 microglia: by construction, only the top 5 % of 4s2 cells exceed it. The two callers agree on the majority of cells; we keep the raw-count call downstream because integer counts are easier to interpret and do not depend on the `AddModuleScore()` binning parameters.

### Apply the call and cross-check

```{r}
#| label: switch-call

# Raw-count call: switch+ if the cell carries at least K_raw reads across the two E2 probes.
mg$apoe_switch <- factor(if_else(mg$e2_sum >= K_raw, "switch+", "switch-"),
                         levels = c("switch+", "switch-"))
# Module-score call: kept only for the cross-tab below.
mg$apoe_switch_mod <- factor(if_else(mg$apoe_E2_module1 >= T_module, "switch+", "switch-"),
                             levels = c("switch+", "switch-"))

# Per-condition 2 x 2 of the two callers; perfect agreement would put all mass on the diagonal.
# Qualify dplyr::count because matrixStats::count is loaded via DESeq2 and masks it.
mg@meta.data |>
  as_tibble() |>
  dplyr::count(condition, apoe_switch, apoe_switch_mod) |>
  pivot_wider(names_from = apoe_switch_mod, values_from = n, values_fill = 0L) |>
  arrange(condition, apoe_switch) |>
  gt(groupname_col = "condition", rowname_col = "apoe_switch") |>
  tab_header(title    = "Raw-count vs module-score switch call",
             subtitle = md("Rows: raw-count call (K_raw)<br>columns: T-module-score call (95th pct of 4s2)")) |>
  cols_label(`switch+` = "module: switch+",
             `switch-` = "module: switch-") |>
  tab_options(table.font.size = px(12))
```

The two callers agree on the bulk of cells. We use `apoe_switch` (raw count) downstream.

## Per-sample switch efficiency

In this secton we investigate the percentage of cells expressing the E2 transcript out of the total number of cells identified as microglia in each sample.

```{r}
#| label: switch-by-sample
#| fig-width: 9
#| fig-height: 4

# Per-sample switch fraction; this is the answer to "50 out of 200 microglia in sample x express E2".
sample_switch <- mg@meta.data |>
  as_tibble() |>
  summarize(n_microglia    = n(),
            n_switchplus   = sum(apoe_switch == "switch+"),
            pct_switchplus = 100 * n_switchplus / n_microglia,
            .by = c(sample_id, condition)) |>
  arrange(condition, sample_id)

sample_switch |>
  gt(rowname_col = "sample_id") |>
  fmt_number(columns = pct_switchplus, decimals = 1) |>
  tab_style(style    = cell_text(weight = "bold"),
            locations = cells_body(columns = pct_switchplus,
                                   rows    = condition == "switched" & pct_switchplus < 30)) |>
  tab_header(title    = "Microglia switch+ (%) by sample") |>
  tab_options(table.font.size = px(12))

# Bar chart with both reference lines: 5 % is the FP baseline (by construction on 4s2), 30 % is the
# soft floor for a "biologically switched" 4s2M sample.
sample_switch |>
  ggplot(aes(sample_id, pct_switchplus, fill = condition)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 5,  linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = c(control = "firebrick", switched = "steelblue")) +
  labs(x = NULL, y = "% switch+ microglia",
       title    = "Switch efficiency per sample",
       subtitle = "Dashed = 5 % FP baseline (4s2 by construction)") +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "bottom")
```



**Microglia-purity caveat.** These percentages are only as clean as the microglia object itself.  
Any cell from another lineage (oligodendrocyte, astrocyte, neuron, T cell, vascular) that segmented as microglia and survived the contamination cleaning in `20260517_microglia_subcluster.qmd` will dilute the switch+ fraction in the 4s2M arm and slightly inflate the 4s2 baseline.  


## UMAP and spatial QC of the switch call

```{r}
#| label: switch-umap
#| fig-width: 12
#| fig-height: 5
#| column: page
#| eval: FALSE



# A clean switch call should sprinkle switch+ cells across all microglia states, not concentrate them
# in one cluster; the latter would mean the score is tracking a transcriptional axis, not genotype.
DimPlot(mg,
        reduction = "umap",
        group.by  = "apoe_switch",
        split.by  = "condition",
        cols      = c(`switch+` = "red", `switch-` = "grey97"),
        order     = "switch+",
        pt.size   = 0.5) +
  ggtitle("Microglia UMAP coloured by switch call")
```

```{r}
#| label: switch-spatial
#| fig-width: 14
#| fig-height: 7
#| column: page

# The microglia object may have been DietSeurat'd without FOVs; if so, lift the call onto the parent
# SpaNorm object (which still carries fov / fov.2) and plot from there.
if (length(Images(mg)) > 0) {
  spatial_obj <- mg
} else {
  spatial_obj <- qs_read("data/20260516_microglia_switch_spanorm_p025_clean.qs2")
  # Direct meta.data assignment by row name; `obj$col[cells] <- ...` does not write back through the
  # Seurat $<- accessor, so we touch the data.frame directly and then refactor.
  spatial_obj@meta.data$apoe_switch <- NA_character_
  spatial_obj@meta.data[colnames(mg), "apoe_switch"] <- as.character(mg$apoe_switch)
  spatial_obj$apoe_switch <- factor(spatial_obj$apoe_switch,
                                    levels = c("switch+", "switch-"))
}

# Cell filter restricts the panel to microglia (the only cells with a non-NA call); white background
# keeps tissue edges readable.
ImageDimPlot(spatial_obj,
             fov             = c("fov", "fov.2"),
             cells           = colnames(spatial_obj)[!is.na(spatial_obj$apoe_switch)],
             group.by        = "apoe_switch",
             cols            = c(`switch+` = "red", `switch-` = "grey90"),
             size            = 0.7,
             dark.background = FALSE) +
  plot_annotation(title = "switch+ vs switch- microglia in situ")
```

## Pseudobulk DE on the switch call

### Switch+ vs switch- cells _within_ 4s2M:

Each 4s2M sample contributes up to two pseudobulks (one switch+, one switch-)

```{r}
#| label: pb-switch-within
#| fig-width: 13
#| fig-height: 6

# Subset to the switched arm; each 4s2M sample contributes up to two pseudobulks (one switch+, one switch-).
sub_within <- subset(mg, condition == "switched")
# AggregateExpression mangles underscores into hyphens and prepends 'g' to digit-leading strings, so we
# encode switch+/- as swP / swN to keep the suffix recoverable from the mangled column names.
sub_within$bulk_label <- paste(sub_within$sample_id,
                               ifelse(sub_within$apoe_switch == "switch+", "swP", "swN"),
                               sep = "_")

bulk_within <- AggregateExpression(sub_within,
                                   assays        = "Xenium",
                                   group.by      = "bulk_label",
                                   return.seurat = TRUE)

# Recover switch_call from the mangled suffix; "-swP" / "-swN" survive the underscore-to-hyphen rewrite.
bulk_within$switch_call <- if_else(str_ends(colnames(bulk_within), "-swP"),
                                   "switch+", "switch-")
Idents(bulk_within) <- "switch_call"

res_within <- anayet(bulk_within,
                     ident1          = "switch+",
                     ident2          = "switch-",
                     test_method     = "DESeq2",
                     p_adj_threshold = 0.1,
                     min_pct         = 0.3,
                     plot_title      = "4s2M microglia DE",
                     print_plot      = FALSE)

# Side-by-side volcano + top-hit table, same idiom as the per-state loop in
# 20260518_microglia_pseudobulk_DE.qmd; falls back to top-5-by-raw-p when no gene clears FDR < 0.1.
top_within <- if (nrow(res_within$significant_results) > 0) {
  res_within$significant_results |>
    as_tibble(rownames = "gene") |>
    select(gene, p_val, avg_log2FC, p_val_adj) |>
    mutate(across(where(is.numeric), \(x) signif(x, 3))) |>
    arrange(p_val_adj)
} else {
  res_within$all_results |>
    as_tibble(rownames = "gene") |>
    slice_min(p_val, n = 5, with_ties = FALSE) |>
    select(gene, p_val, avg_log2FC, p_val_adj) |>
    mutate(across(where(is.numeric), \(x) signif(x, 3)))
}
tbl_within <- tableGrob(top_within, theme = ttheme_minimal(base_size = 11), rows = NULL)

res_within$volcano_plot + wrap_elements(tbl_within) + plot_layout(widths = c(3, 4))
```



### E4 vs E2 sensitivity, restricted to switch+ in 4s2M

All cells in control mice (4s2) microglia vs. cells called as (switch +) in 4s2M mice microglia
```{r}
#| label: pb-e2-restricted
#| fig-width: 13
#| fig-height: 6

# 4s2 cells stay in full; 4s2M cells are restricted to switch+ so the "switched" arm is enriched for
# cells that actually carry E2 transcript.
sub_restr <- subset(mg, condition == "control" |
                        (condition == "switched" & apoe_switch == "switch+"))

bulk_restr <- AggregateExpression(sub_restr,
                                  assays        = "Xenium",
                                  group.by      = "sample_id",
                                  return.seurat = TRUE)

# Mirror the metadata-join idiom from the parent pseudobulk qmd: AggregateExpression rewrites
# "4s2_F1" -> "g4s2-F1", so we reconstruct the lookup with the same transform.
sample_meta <- sub_restr@meta.data |>
  distinct(sample_id, condition) |>
  mutate(bulk_id = paste0("g", gsub("_", "-", sample_id)))

bulk_restr$condition <- sample_meta$condition[match(colnames(bulk_restr), sample_meta$bulk_id)]
stopifnot(!any(is.na(bulk_restr$condition)))
Idents(bulk_restr) <- "condition"

res_restr <- anayet(bulk_restr,
                    ident1          = "switched",
                    ident2          = "control",
                    test_method     = "DESeq2",
                    p_adj_threshold = 0.1,
                    min_pct         = 0.3,
                    plot_title      = "Restricted-to-switch+ DE",
                    print_plot      = FALSE)

top_restr <- if (nrow(res_restr$significant_results) > 0) {
  res_restr$significant_results |>
    as_tibble(rownames = "gene") |>
    select(gene, p_val, avg_log2FC, p_val_adj) |>
    mutate(across(where(is.numeric), \(x) signif(x, 3))) |>
    arrange(p_val_adj)
} else {
  res_restr$all_results |>
    as_tibble(rownames = "gene") |>
    slice_min(p_val, n = 5, with_ties = FALSE) |>
    select(gene, p_val, avg_log2FC, p_val_adj) |>
    mutate(across(where(is.numeric), \(x) signif(x, 3)))
}
tbl_restr <- tableGrob(top_restr, theme = ttheme_minimal(base_size = 11), rows = NULL)

res_restr$volcano_plot + wrap_elements(tbl_restr) + plot_layout(widths = c(3, 4))
```



```{r}
#| label: pb-save
#| eval: false

# Persist both pseudobulk tables; rerun this chunk only after the upstream calibration has changed.
res_within$all_results |>
  as_tibble(rownames = "gene") |>
  write_csv("results/20260523-pseudobulk_DE_microglia_switchplus_vs_switchminus.csv")

res_restr$all_results |>
  as_tibble(rownames = "gene") |>
  write_csv("results/20260523-pseudobulk_DE_microglia_E4vsE2_switchplus_only.csv")
```

## Caveat

- **Only two of the four panel probes carry usable E2 evidence.** `rs429358-ALT-228:C` is insensitive in this assay and `rs7412-WT-226` does not separate arms; both are excluded from the module score. `rs7412-ALT-226:T` is cross-reactive but its delta between conditions still carries E2 information once combined with `rs429358-WT-228`.
- **The switch call inherits the 4s2 noise floor.** `K_raw` is calibrated to keep the 4s2 false-positive rate at 5 %, so by construction roughly one in twenty 4s2 microglia gets called switch+. A 4s2M sample whose switch+ fraction sits close to that 5 % baseline failed to recombine and should be treated as biologically equivalent to a control mouse.



## References

- Parent pseudobulk DE notebook: `20260518_microglia_pseudobulk_DE.qmd`.
- 10x Genomics Xenium custom panel design documentation (panel: mBrain_480g).

