---
title: "SpaNorm: spatial-aware normalization"
subtitle: "sample.p = 0.25, (package default)"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
---

## Why this notebook

The primary SpaNorm notebook (`20260512_SpaNorm.qmd`) runs at `sample.p = 0.05`, the value Salim et al. (Genome Biology, 2025) recommend for ~400k-cell Xenium datasets. The package default is 0.25. We re-run the same pipeline at the default here to test whether the cluster structure is stable under a five-fold larger fitting subsample. If the per-slide and per-sample UMAPs match, we keep `sample.p = 0.05` for the project; if they diverge, we promote 0.25.

Every other parameter, the per-sample fitting strategy, and the downstream `ScaleData -> PCA(30) -> FindNeighbors -> Louvain(0.5) -> UMAP` chain stay identical so the comparison is clean.

Bibliography for this notebook:

- Salim et al., "SpaNorm: spatially-aware normalization for spatial transcriptomics data," Genome Biology, 2025. DOI: 10.1186/s13059-025-03565-y.
- SpaNorm package vignette: https://bhuvad.github.io/SpaNorm/articles/SpaNorm.html.
- Cached NotebookLM read-through: `notebookLM/2026-05-12_spanorm-background.md`.

## Setup

```{r}
#| label: setup

# Seurat v5 holds the merged Xenium object with the per-slide FOV slots we read centroids from.
library(Seurat)
library(tidyverse)
# qs2 reads the ~5 GB checkpoint roughly 10x faster than saveRDS.
library(qs2)
library(fs)
# SpaNorm is the spatially aware normalization (Salim, Genome Biology, 2025).
library(SpaNorm)
# SpatialExperiment is the native input class for SpaNorm; we build one per sample.
library(SpatialExperiment)
# scCustomize only contributes its categorical palette helper for the UMAP plots.
library(scCustomize)

# Seed Louvain and any other stochastic step so cluster labels stay stable across renders.
set.seed(1)

# Reload the post-SpaNorm checkpoint; built by the chunk below the first time it is run with eval:true.
xen <- qs_read("data/20260512_microglia_switch_spanorm_p025.qs2")
```

## Build the SpaNorm assay (one fit per sample, sample.p = 0.25)

```{r}
#| label: spanorm
#| eval: false
# eval:false because this is the one-off build; every render reloads the checkpoint above.

# Sample, slide, and condition lookup; matches R/20260512_xenium2seurat.R for self-containment.
samples <- tribble(
  ~sample_id, ~slide,            ~condition,
  "4s2_F1",   "Slide1_0022474",  "control",
  "4s2M_F1",  "Slide1_0022474",  "switched",
  "4s2M_F2",  "Slide1_0022474",  "switched",
  "4s2_F2",   "Slide2_0069080",  "control",
  "4s2_F3",   "Slide2_0069080",  "control",
  "4s2M_F3",  "Slide2_0069080",  "switched",
  "4s2M_F4",  "Slide2_0069080",  "switched")

# Load the raw merged object; SpaNorm needs raw counts, not the SCT residuals.
xen <- qs_read("data/20260512_microglia_switch_original_seurat.qs2")
DefaultAssay(xen) <- "Xenium"

# Build the sample_id -> FOV-name lookup from the object itself; LoadXenium uses default name "fov" and merge renames collisions
# (so the actual names are usually fov / fov.2, not the slide labels).
fov_to_slide <- map_chr(Images(xen), \(img) {
  # Each FOV's cells share the same slide value in meta.data; use the first one as the slide tag.
  unique(xen@meta.data[Cells(xen[[img]]), "slide"])[1]
}) |> set_names(Images(xen))
# Invert: slide -> FOV, then index by sample slide to get the per-sample FOV lookup.
slide_to_fov <- set_names(names(fov_to_slide), unname(fov_to_slide))
fov_lookup <- set_names(slide_to_fov[samples$slide], samples$sample_id)

# Run SpaNorm on one sample at the package-default subsample. Returns the log-scale normalized matrix.
spanorm_one_sample <- function(s, xen, fov_lookup) {
  # Subset keeps only the cells of one tissue section; SpaNorm's smoothness assumption is per section.
  seu_s <- subset(xen, subset = sample_id == s)
  # Centroids come from the slide FOV; we filter to the cells that survived the per-sample subset.
  coords <- GetTissueCoordinates(seu_s, image = fov_lookup[[s]]) |>
    filter(cell %in% Cells(seu_s))
  # SpatialExperiment is SpaNorm's native input; we hand it raw counts plus the 2D centroids.
  spe <- SpatialExperiment(
    assays        = list(counts = GetAssayData(seu_s, assay = "Xenium", layer = "counts")),
    spatialCoords = as.matrix(coords[, c("x", "y")]))
  # sample.p=0.25 is the package default; the only change from 20260512_SpaNorm.qmd.
  spe <- SpaNorm(spe,
                 sample.p   = 0.25,
                 df.tps     = 6,
                 adj.method = "auto",
                 backend    = "auto",
                 verbose    = TRUE)
  # logcounts holds the normalized matrix; we keep Seurat's cell barcodes for the later cbind.
  out <- SummarizedExperiment::assay(spe, "logcounts")
  colnames(out) <- Cells(seu_s)
  out}

# Loop over the 7 samples; map keeps the per-sample matrices named for the reduce step.
norm_list <- map(samples$sample_id, spanorm_one_sample,
                 xen = xen, fov_lookup = fov_lookup) |>
  set_names(samples$sample_id)

#qd_save(norm_list, "data/20260513_norm_list_sample025.qs2")

# Bind the per-sample matrices into one wide normalized matrix and reorder to the merged object's cell order.
# Qualify reduce() because SpatialExperiment loads IRanges, whose reduce() masks purrr::reduce.
norm_mat <- purrr::reduce(norm_list, cbind)
norm_mat <- norm_mat[, Cells(xen)]

# Attach as a new v5 assay; data layer only because the auto-selected adjustment returns log-scale residuals.
# CreateAssay5Object avoids the "counts layer not present" warning that CreateAssayObject (v3) triggers inside a v5 object.
xen[["SpaNorm"]] <- CreateAssay5Object(data = norm_mat)
DefaultAssay(xen) <- "SpaNorm"

# Use all 480 genes as variable features; on a targeted Xenium panel every gene was chosen as informative by design,
# and FindVariableFeatures("vst") requires a counts layer the SpaNorm assay does not carry.
VariableFeatures(xen) <- rownames(xen)

# RunPCA expects a scale.data layer; SCT writes it implicitly, but our custom SpaNorm assay only carries the data layer.
xen <- ScaleData(xen, verbose = FALSE)

# Default npcs (50) is fine; we only use the first 30 downstream.
xen <- RunPCA(xen, npcs = 30, verbose = FALSE)

# 1:30 PCs and resolution 0.5 match 20260512_SpaNorm.qmd for a like-for-like comparison.
xen <- FindNeighbors(xen, reduction = "pca", dims = 1:30, verbose = FALSE)
xen <- FindClusters(xen, resolution = 0.5, algorithm = 1, verbose = FALSE)
xen <- RunUMAP(xen, reduction = "pca", dims = 1:30, verbose = FALSE)


# qs_save(xen, "data/20260512_microglia_switch_spanorm_p025.qs2")
```

## PCA split by slide

```{r}
#| label: palette
# Palette sized to the actual cluster count so colors stay stable if the resolution changes the number of clusters.
colorines <- scCustomize_Palette(num_groups = n_distinct(xen$seurat_clusters), ggplot_default_colors = FALSE)
```

```{r}
#| label: pca
# Split by slide to compare against the sample.p=0.05 PCA in 20260512_SpaNorm.qmd.
PCAPlot(xen, split.by = "slide", cols = colorines)
```

**Result.** The PC1/PC2 envelope at `sample.p = 0.25` reproduces the fan shape from `20260512_SpaNorm_sample005.html`, with Slide1 and Slide2 occupying matched coordinate ranges (PC1 from -10 to ~15; PC2 from -8 to ~28).  

Cluster identities differ between the two builds (cluster 4, green, dominates the upper PC2 region here, whereas cluster 2, red, did so at `sample.p = 0.05`); that is the expected Louvain re-labelling under slightly shifted PCA coordinates. The fan structure itself is stable. We read this as PCA-level stability under a five-fold larger fitting subsample.

## UMAP

```{r}
#| label: umap
DimPlot(xen, cols = colorines)
```

**Result.** `sample.p = 0.25` yields 25 clusters, the same count as `sample.p = 0.05` at the matched Louvain resolution.  

 The 2D layout re-orients (cluster 0, dark, now sits at the top of the UMAP rather than at the bottom-center), but the same set of discrete islands appears, including the small satellite clusters near the right edge of the embedding (clusters 14 and 17 at umap_1 ~ 14). Cluster count and topology are stable; the absolute UMAP layout is not, which is the expected behavior under any seeded re-embedding.

### Per slide

```{r}
#| label: umap-slide
#| fig-width: 10
#| column: page

# Side-by-side UMAPs per slide; the failure mode we test against is any cluster that only exists in one slide.
DimPlot(xen, split.by = "slide", cols = colorines) + NoLegend()
```

**Result.**  The two slide-imbalance regions flagged at `sample.p = 0.05` (top-left island carrying different cluster identities, bottom-center lobe denser on Slide2) are much less pronounced at `sample.p = 0.25`. 

Both slides now populate the same cluster geometry with comparable densities: the dark cluster 0 at the top, the gray/red lower-left lobe, and the right-edge satellites all replicate on both panels. The package default absorbs the slide effect more cleanly than the 5 % subsample on this dataset.   

On the strength of this panel alone, we promote `sample.p = 0.25` to the project default and revisit the need for Harmony only if the three-way normalization comparison still shows residual slide structure.

### Per sample

```{r}
#| label: umap-sample
#| fig-width: 14
#| fig-height: 12
#| column: page

# Per-sample facets; 4s2M_F2 (partial section) is the sample most likely to behave differently.
DimPlot(xen, split.by = "sample_id", ncol = 3, cols = colorines) + NoLegend()
```

**Result.** All six complete sections (4s2_F1, F2, F3 and 4s2M_F1, F3, F4) reproduce the 25-cluster landscape with similar proportions.   

4s2M_F2 still contributes fewer cells, as expected for the partial section, but the cells it does contribute land in the same basins as the matched complete sections, with cluster 0 (dark) at the top and the lower-left gray/red lobe both well represented. The partial-section flag stays a cell-count flag, not a normalization or cluster-identity flag.

## Summary

Per-sample SpaNorm at `sample.p = 0.25` produces the same 25-cluster decomposition as `sample.p = 0.05` and better balances the two slides on the per-slide UMAP. Convergence is clean on all seven fits (see `20260513_SpaNorm_samplep025_convergence.md`). We adopt `sample.p = 0.25` as the project default and update the SCT-vs-SpaNorm three-way comparison to read this build's checkpoint.

## References

- Salim A et al. "SpaNorm: spatially-aware normalization for spatial transcriptomics data." Genome Biology, 2025. DOI: 10.1186/s13059-025-03565-y.
- SpaNorm vignette: https://bhuvad.github.io/SpaNorm/articles/SpaNorm.html.
- Cached NotebookLM read-through: `notebookLM/2026-05-12_spanorm-background.md`.
